---
title: "DataViz for SocScientists Notes"
author: "Caitlin S. Ducate"
date: "5/21/2020"
output: html_document
---

```{r, eval = FALSE}
# Setup
myPackages <- c("tidyverse", "broom", "coefplot", "cowplot",
                 "gapminder", "GGally", "ggrepel", "ggridges", "gridExtra",
                 "here", "interplot", "margins", "maps", "mapproj",
                 "mapdata", "MASS", "quantreg", "rlang", "scales",
                 "survey", "srvyr", "viridis", "viridisLite", "devtools")
install.packages(myPackages)
devtools::install_github("kjhealy/socviz")

```

## Chapter 1: Look At Data

**Why look at data?**

* Because numbers can be misleading & describe a variety of patterns that will only come to light when we can see all of the data at once

**Principles of bad figure making**

* "Chart junk": extraneous stuff that doesn't add to the data story
  + In some cases, though, a memorable graph will have a bit of superfluous design if it is clever
* Bad data: the data being presented tell a misleading story
* Problems with perception: the chart may be free of junk, but human visual perception will be misled by the chart's layout or dimensions

**Human Perception**

* Humans are better at seeing gradients when they are all the same hue and chroma but vary in luminance
* Need to be careful with color choice to make sure colors step through the options as intended
  + In other words, colors can be misleading if picked wrong (e.g. one color can unintentionally stand out more than the others)
* Shape and color are two "channels" that can encode information visually about your data
  + Color channel seems to work better than shape channel
  + Should try to avoid showing data through multiple channels
* Gestalt Rules
  + Proximity: things close together seem related
  + Similarity: things that look alike seem related
  + Connection: things visually tied together seem related
  + Continuity: Partially hidden objects are perceptually completed
  + Figure & ground: visual elements seen in either the foreground or the background
  + Common fate: elements moving in the same direction are seen as a unit (e.g. school of fish)

**Decoding Graphs**

* Humans do best when judging the relative position of things on a common scale
* Humans do worst when judging quantities as angles or areas (esp. areas of circles)

**Honest & Good Judgment** 

* Not always good rules of thumb for what is an honest representation
  + Sometimes it makes sense not to start your Y-axis at 0, and if your axes are labeled, not necessarily misleading
<<<<<<< HEAD
  
## Chapter 2: Getting Started

```{r, results=FALSE}
# Load in libraries
library(tidyverse)
library(socviz)
```

* Mostly an overview of R & RStudio
* A `tibble` is a tidyverse data.frame

```{r}
# Tiny data set from socviz package
class(titanic)

# Turn titanic into a tidyverse tibble
titanic_tb <- as_tibble(titanic)
titanic_tb
```

* The package `haven` is also good for reading in data of various formats
* Apparently "tidy" data is long format rather than wide format
  + **Note**: I am intrigued

### Making my first figure

```{r}
library(gapminder)
gapminder

p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point()
```

